Spherical harmonic analysis of faceted spheroids identifies shaping strategies and standardisation at Qianshangying (North China)

### 2025-08-12 analysis/paper/paper.qmd
working on figures

lava_mluti_text <- df_long %>% filter(material == "Lava", category == "Polys_and_Multi_cores") %>% pull(text)
dolomite_mluti_text <- df_long %>% filter(material == "Dolomite", category == "Polys_and_Multi_cores") %>% pull(text)
chert_mluti_text <- df_long %>% filter(material == "Chert", category == "Polys_and_Multi_cores") %>% pull(text)
other_mluti_text <- df_long %>% filter(material == "Other", category == "Polys_and_Multi_cores") %>% pull(text)
```

@@

```{r}
#| label: fig-basic-characteristics
#| fig-cap: "Basic characteristics of the Qianshangying assemblage. A: Raw material proportions. B: Box plot comparing of core, spheroid, and multifacial core dimensions"

@@

ggplot(data_clean, aes(x = Sphericity, y = `SPHARM power`)) +
  geom_point(size = 3, alpha = 0.9, color = "#F28E2B") +
  geom_smooth(method = "lm", se = TRUE, color = "#1C3041", fill = "#A6CEE3") +
  scale_x_continuous(breaks = seq(0.4, 1.0, by = 0.04)) +
  theme_minimal(base_size = 15) +
  theme(
    panel.grid.major = element_line(color = "#DDDDDD"),
    panel.grid.minor = element_blank(),
    axis.title = element_text(color = "#333333"),
    axis.text = element_text(color = "#333333")
  ) +
  labs(x = "Sphericity", y = "SHE") +
  annotate("text",
         x = Inf, y = Inf,
         hjust = 1.1, vjust = 1.2,
         label = paste0("r = ", round(correlation, 3),
                        "\n", "p = ", pretty_print_sci(p_value),
                        "\n", "n = ", nrow(data_clean)),
         size = 5, color = "#1C3041")
```

@@

# 自定义Group名称映射，方便图例显示
df_long <- df_long %>%
  mutate(Group = factor(Group, levels = c("Spheroid + Subspheroid", "Polyhedron + Multifacial"),
                        labels = c("Spheroid + Subspheroid", "Polyhedron + Multifacial")))

# 画图
ggplot(df_long, aes(x = Group, y = Value, fill = Group)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA, width = 0.5) +
  geom_jitter(width = 0.15, alpha = 1, size = 1) +
  facet_wrap(~ Variable, scales = "free_y") +
  scale_fill_manual(values = c("Spheroid + Subspheroid" = "orange", "Polyhedron + Multifacial" = "#66c2a5")) +
  labs(x = "", y = "Value", fill = "Group", title = "Sphericity and SHE by Group") +
  theme_minimal(base_size = 10) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 0, hjust = 0.5)
  )
```

@@

ggplot(df, aes(x = degree, y = variance)) +
  geom_line(color = "#1f78b4", size = 1.3, alpha = 0.8) +          
  geom_point(color = "#ff7f00", size = 3, alpha = 0.9, shape = 21, fill = "#ffbb78", stroke = 1) +  # 橙色边框，填充浅橙
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) +    
  scale_y_continuous(breaks = scales::pretty_breaks(n = 8)) +      
  labs(
    title = "Variance per Degree",
    x = "Spherical Harmonic Degree",
    y = "Variance"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(face = "bold"),
    panel.grid.major = element_line(color = "#DDDDDD", size = 0.5),
    panel.grid.minor = element_blank()
  )

@@

@@

# 绘图，带误差带
ggplot(summary_df, aes(x = degree, y = mean_power, color = group, fill = group)) +
  geom_line(linewidth = 1) +
  geom_ribbon(aes(
    ymin = pmax(mean_power - sd_power, 1e-10),
    ymax = mean_power + sd_power
  ), alpha = 0.2, linetype = 0) +
  scale_y_log10(
  breaks = c(1e-8, 1e-6, 1e-4, 1e-2, 1),
  labels = parse(text = c("10^-8", "10^-6", "10^-4", "10^-2", "10^0"))
  )+
  scale_x_continuous(
    limits = c(0, 20),
    breaks = seq(0, 20, by = 4),
    expand = c(0, 0)
  ) +
  coord_cartesian(clip = "on") +
  scale_color_manual(values = c(
    "All specimens" = "lightgray",
    "Spheroid/Subspheroid" = "orange",
    "Polyhedron/Multifacial" = "#66c2a5"
  )) +
  scale_fill_manual(values = c(
    "All specimens" = "lightgray",
    "Spheroid/Subspheroid" = "orange",
    "Polyhedron/Multifacial" = "#66c2a5"
  )) +
  labs(
    title = "Mean Power Spectrum with Standard Deviation",
    x = "Degree (l)",
    y = "Mean Total Power (log scale)",
    color = "Group",
    fill = "Group"
  ) +
  theme_minimal() +
  theme(plot.margin = margin(t = 10, r = 10, b = 10, l = 10),
  legend.position = c(0.95, 0.95),
  legend.justification = c("right", "top"),
  legend.background = element_rect(fill = alpha("white", 0.8), color = NA),
  legend.box.background = element_rect(color = "gray80"),
  legend.box.margin = margin(5, 5, 5, 5)
  )
```

```{r, include=FALSE}
library(readxl)
library(dplyr)
library(ggplot2)
library(tidyr)

# 读取 Excel 文件
file_path <- "E:/Lithic Analysis/QSYSpheroidsStudy/analysis/data/derived_data/umap_l2_l3.XLSX"
df <- read_excel(file_path)

# 查看数据结构
print(head(df))

# 计算相关系数矩阵
correlations <- df %>%
  select(UMPA1, UMAP2, degree1, degree2) %>%
  cor()

print("相关系数矩阵：")
print(correlations)

# 将数据长格式化方便绘图
df_long <- df %>%
  pivot_longer(cols = c(degree1, degree2), names_to = "degree", values_to = "degree_value")

# 绘制 UMAP1 和 degree的散点图
p1 <- ggplot(df_long, aes(x = UMPA1, y = degree_value, color = degree)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "UMAP1 vs degree1 (l=2) and degree2 (l=3)",
       x = "UMAP1",
       y = "Degree Power") +
  theme_minimal()

# 绘制 UMAP2 和 degree的散点图
p2 <- ggplot(df_long, aes(x = UMAP2, y = degree_value, color = degree)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "UMAP2 vs degree1 (l=2) and degree2 (l=3)",
       x = "UMAP2",
       y = "Degree Power") +
  theme_minimal()

print(p1)
print(p2)

```
We calculated the spherical sphericity ans spherical harmonic energy (SHE) to both the spheroids (n = `r n_total_typology`) and all other multifacial and polyhedron cores (n = `r n_poly_multi`) in the Qianshangying assemblage. These analyses aim to assess whether spheroids are morphologically distinct from other cores, and whether they exhibit standardized shapes—an essential criterion for evaluating the existence of deliberate shaping strategies and a goal product. In addition, we compare quantitative outcomes with traditional typological classifications to assess their consistency. We present a summary table listing each core's blank morphology and type, final morphology, and cortex coverage here in [@tbl-1].

>>>>

lava_multi_text <- df_long %>% filter(material == "Lava", category == "Polys_and_Multi_cores") %>% pull(text)
dolomite_multi_text <- df_long %>% filter(material == "Dolomite", category == "Polys_and_Multi_cores") %>% pull(text)
chert_multi_text <- df_long %>% filter(material == "Chert", category == "Polys_and_Multi_cores") %>% pull(text)
other_multi_text <- df_long %>% filter(material == "Other", category == "Polys_and_Multi_cores") %>% pull(text)
```

@@

```{r}
#| label: fig-basic-characteristics
#| fig-cap: "Basic characteristics of the Qianshangying assemblage. A: Raw material proportions. B: Box plot comparing of core, spheroid, and multifacial core dimensions"
#| echo: false

@@

p_corr_sphericiy_she <- 
ggplot(data_clean, aes(x = Sphericity, y = `SPHARM power`)) +
  geom_point(size = 3, alpha = 0.9, color = "grey60") +
  geom_smooth(method = "lm", se = TRUE, color = "black", fill = "#A6CEE3") +
  scale_x_continuous(breaks = seq(0.4, 1.0, by = 0.04)) +
  theme_minimal(base_size = 12)  +
  labs(x = "Sphericity", 
       y = "Spherical Harmonic Energy") +
  annotate("text",
         x = Inf, y = Inf,
         hjust = 1.1, vjust = 1.2,
         label = paste0("r = ", round(correlation, 3),
                        "\n", "p = ", pretty_print_sci(p_value),
                        "\n", "n = ", nrow(data_clean)),
         size = 5, color = "#1C3041")
```

@@

# 自定义Group名称映射，方便图例显示
df_long <- df_long %>%
  mutate(Group = factor(Group, 
                        levels = c("Spheroid + Subspheroid", 
                                   "Polyhedron + Multifacial"),
                        labels = c("Spheroid + Subspheroid", 
                                   "Polyhedron + Multifacial"))) %>% 
  mutate(
    Group = case_when(
      Group == "Spheroid + Subspheroid" ~  "Spheroid\nSubspheroid", 
      Group == "Polyhedron + Multifacial" ~  "Polyhedron\nMultifacial",
      TRUE ~ NA_character_
    )
  )

# 画图
p_sphericity_and_she <- 
ggplot(df_long, aes(x = Group, 
                    y = Value, 
                    fill = Group,
                    colour = Group)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA, width = 0.5) +
  geom_jitter(width = 0.15, 
              alpha = 1, 
              size = 1) +
  facet_wrap(~ Variable, scales = "free_y") +
  scale_fill_manual(values = c("Spheroid\nSubspheroid" = "orange", 
                               "Polyhedron\nMultifacial" = "#66c2a5")) +
  scale_colour_manual(values = c("Spheroid\nSubspheroid" = "orange", 
                               "Polyhedron\nMultifacial" = "#66c2a5")) +
  labs(x = "",
       y = "",
       fill = "Group", 
      # title = "Sphericity and SHE by Group"
       ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none"
  )
```

@@

p_variance_per_degreee <- 
ggplot(df, aes(x = degree, y = variance)) +
  geom_line(size = 1.3, colour = "black") +          
  geom_point(size = 3, colour = "grey60") +  # 橙色边框，填充浅橙
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) +    
  scale_y_continuous(breaks = scales::pretty_breaks(n = 8)) +      
  labs(
   # title = "Variance per Degree",
    x = "Spherical Harmonic Degree",
    y = "Variance"
  ) +
  theme_minimal(base_size = 12)

@@

# 绘图，带误差带
p_power_spectrum <- 
ggplot(summary_df, aes(x = degree, y = mean_power, color = group, fill = group)) +
  geom_line(linewidth = 1) +
  geom_ribbon(aes(
    ymin = pmax(mean_power - sd_power, 1e-10),
    ymax = mean_power + sd_power
  ), alpha = 0.2, linetype = 0) +
  scale_y_log10(
  breaks = c(1e-8, 1e-6, 1e-4, 1e-2, 1),
  labels = parse(text = c("10^-8", "10^-6", "10^-4", "10^-2", "10^0"))
  )+
  scale_x_continuous(
    limits = c(0, 20),
    breaks = seq(0, 20, by = 4),
    expand = c(0, 0)
  ) +
  coord_cartesian(clip = "on") +
  scale_color_manual(values = c(
    "All specimens" = "lightgray",
    "Spheroid/Subspheroid" = "orange",
    "Polyhedron/Multifacial" = "#66c2a5"
  )) +
  scale_fill_manual(values = c(
    "All specimens" = "lightgray",
    "Spheroid/Subspheroid" = "orange",
    "Polyhedron/Multifacial" = "#66c2a5"
  )) +
  labs(
   # title = "Mean Power Spectrum with Standard Deviation",
    x = "Degree (l)",
    y = "Mean Total Power (log scale)",
    color = "Group",
    fill = "Group"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.margin = margin(t = 10, r = 10, b = 10, l = 10),
  legend.position = c(0.95, 0.4),
  legend.justification = c("right", "top"),
  legend.background = element_rect(fill = alpha("white", 0.8), color = NA),
  legend.box.background = element_rect(color = "gray80"),
  legend.box.margin = margin(5, 5, 5, 5)
  )
```

```{r}
#| label: fig-sp-summary
#| fig-cap: "A. Mean spherical harmonic power spectrum. B. Variance distribution by spherical harmonic degree. C. Distribution of SPHARM power and sphericity by artefact type. D. Correlation of sphericity and spherical harmonic energy"

# combine plots
library(cowplot)
p_panel_spharm <- 
plot_grid(p_power_spectrum, 
          p_variance_per_degreee,
          p_sphericity_and_she,
          p_corr_sphericiy_she,
          labels = 'AUTO'
           )

ggsave(plot = p_panel_spharm,
       filename = here("analysis/paper/figures/panel_spharm_plots.png"),
       h = 8, 
       w = 9)

knitr::include_graphics(here("analysis/paper/figures/panel_spharm_plots.png"))

@fig-sp-summary

@@

We calculated the spherical sphericity and spherical harmonic energy (SHE) to both the spheroids (n = `r n_total_typology`) and all other multifacial and polyhedron cores (n = `r n_poly_multi`) in the Qianshangying assemblage. These analyses aim to assess whether spheroids are morphologically distinct from other cores, and whether they exhibit standardized shapes—an essential criterion for evaluating the existence of deliberate shaping strategies and a goal product. In addition, we compare quantitative outcomes with traditional typological classifications to assess their consistency. We present a summary table listing each core's blank morphology and type, final morphology, and cortex coverage here in [@tbl-1].
